Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-04 scagnostics.Rmd
First version of discussion.

>>>>

Physics examples motivate us to investigate projection pursuit indexes that detect complex patterns in two dimensional projections. In this context we have investigated the use of indexes developed for variable selection as projection pursuit indexes. Having outlined a set of criteria that define a good index function, we have established methods for the evaluation of the index properties: smoothness, squintability, flexibility, rotation invariance and speed. To this end we used three different sets of simulated data, defined to have different underlying noise distributions (uniform or normal) and different structures (circle, sine and spiral), and we have evaluated several index functions proposed in the literature for variable selection (scagnostics indexes, maximum information based indexes, distance correlation and a spline based index). While all index functions are found to fall short on at least some of the criteria, we have shown that combined with the optimisation available via the guided tour they can already be used to identify interesting views, in particular the built-in structures in the simulated data sets.

We have also sketched some potential improvements of the indexes considered, in particular tuning parameters inherent to the index definition, such as binning, or different smoothing methods for the scagnostics indexes.

Our primary interest is in multidimensional physics models, and we have shown two examples from gravitational wave detection. In the first example, studying a neutron star merger, we have sucessfully used the developped methods to re-discover the non-linear relation between three of the parameters. The second example, stemming from a black hole simulation, has a large parameter space. We used this dataset to explore the results obtained depending on parameter selection, index selection and starting conditions. Interestingly, when working with a small parameter space (six dimensions), projection pursuit uncovers a new relation that has not been anticipated.

We conclude that these methods are promissing and may yield new insight in the inter-dependence of model parameters in the light of experimental data ... other applications can also be envisioned??

There are three frontiers for future work building on our study. First, while we found the considered index functions useful to detect structures, several shortcomings have also been pointed out. While many definitions to identify clusters have been considered, there is much potential for the development of index functions that detect the types of structures considered here. Second, an implementation of interpolations from basis to basis, alleviating the need for rotation invariant indexes is envisioned. Finally, more generally, improved optimisation algorithms are required, e.g. for the identification of structures with small squint angle in high dimensions.
